PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2023-03-10 refereeResponse.txt

Specific comments, notes and suggestions.

TODO: First few edits need to be highlighted in manuscript with strikethroughs and bold face.

Abstract
7: Are short period planets that exotic in the broader context of exoplanets?
	Hard to say! We have replaced 'exotic' with 'wide variety'.
14: "[...] population OF dynamically [...]"
	Fixed
17-18: something is missing in the second part of this sentence.
	Missing words at the end of L17: 'investigate *the effects*'

1.Introduction
26: How large are these solids? I assume this is referencing micron sized dust-grains?
	Correct, some text was added to clarify this.
68: Add reference
	Added a handful of references arguing that the asteroids, KBOs and Oort cloud bodies are leftover planetesimals.
72: Maybe elaborate on this. How long is the period of quiescence? How does this instability work?
	We intentionally left this part vague as someone really needs to look into this more! There are three relevant timescales here:
	1. The timescale for the instability itself (depends on the orbital properties of the embryos, see Chambers 98)
	2. Dynamical friction between protoplanets and planetesimals. This will shut off once the planetesimals are either scattered away, or slowly accreted onto the embryos. Not sure what the timescale for this would be. Note that this shouldn't operate at short period according to this work!
	3. Tidal eccentricity damping of the embryos due to the gas disk. Should shut off once the gas disk disperses on ~Myr timescales (Beckwith + Sargent 1996).
	TODO: decide what to add to the text here.
88: Add reference maybe.
	I don't know of any references which have tested the applicability of statistical methods during the oligarchic growth phase, but maybe some clarification would help. The runaway growth phase produces a small number of large bodies, which eventually interact and terminate runaway growth (Ida + Makino 1993, Kokubo + Ida 1995). Given how few oligarchs are produced, one would not expect particle-in-a-box methods to be a good approach here. We've clarified this in the text as well.
99: Capitalize "Systems of Tightly-packed Inner Planets"
	Fixed
104: Gas-Disk driven migration → reference?
	Added reference to Izidoro 17 and 21, which invokes gas disk driven migration, followed by resonant trapping and an possible orbital instability to explain the observed architectures of STIPs
116: What is the "isolation mass" initial condition? Maybe isolation mass can be defined
	We will define isolation mass later in the paper. For now, we removed this term and replaced it with 'fully-formed, evenly spaced protoplanets'.
118: I'm not sure if this belongs in the introduction. I think the mention of future work at the end of the discussion is sufficient.
	Agreed. Removed the last sentence of this paragraph.

2. Overview of planetesimal accretion
2.1 Oligarchic and Runaway Growth
144 onwards: indices in math upright
	Fixed
152: Define rms velocity
	TODO: not sure what the referee wants? Its defined in equation 1?
161: What does "orderly" fashion mean exactly?
	This means the mass ratios between bodies tend toward unity. This has been clarified in the text.
174: Add reference.
	Added reference to Lissauer 1987. Also mentioned that pebble accretion is a viable alternative.
198: Ln Lambda approx 10 missing could use a reference.
	Typo in the text, Ln Lambda ~ 3 is typically used. Added references to Ida 1990, Stewart and Ida 1993). Ln Lambda = 3 was used to construct figure 1, so no changes to the figure needed.
204: I suggest more formally introducing the concept of Hill sphere and hill factor.
	TODO: Where to put this?
209: This sentence seems not necessary.
	Agreed, removed
213: add "dynamically" heating
	Added
222: add "a" crucial assumption
	Fixed
226: Maybe clarify that growth timescale equals collision time scale.
	We have replaced 'growth' with 'collision' here, and done the same elsewhere in the paper for consistency
238: The eccentricity dispersion e_h is never defined anywhere. It is unclear to me what scaling e_h with the Hill-factor means physically.
	e_h can be thought of as the radial excursion of the body's orbit in units of its Hill radius. For e_h > 1, the random velocity of a body dominates over the velocity difference due to the shearing of the disk (over a separation of 1 r_h) and encounters between bodies can be treated with a two-body formalism. We have decided to remove any mention of e_h here, and instead save this definition for section 4 (TODO: define in section 4), where it is next mentioned.
Figure .1:
Since this is an analytical plot, I think one could color the entire parameter space according to the ratio t_relax/t_coll. Currently, it may suggest that the plotted values for e_h are chosen for a particular reason.
	Get rid of the legend, explain that e_h values increase from bottom to top.
Also, I think it would be nice to have a reference for what values of e_h are standard for certain disk conditions, i.e. how outrageous would it be to have a e_h = 16? Are all disks like that? Or do we expect e_h = 1 as the standard?
	Not sure if there is going to be a reference for this. Equilibrium e_h depends on particle mass and gas density. Could try to plot things in terms of that instead. (Maybe fix rho_gas and vary mass?)
Also, I recommend adjusting the x-limits of the plot to correspond to the limits of the orbital period.
	Done

2.2 Planetesimal Size and Extent of Hill Sphere
253: "will remain proportional" - Proportional to what?
	Changed the wording to 'ratio between the relaxation and collision timescale remains constant' to clarify this.
254: but this doesn't have to be the case at all orbital periods, right? Given Fig.1 it seems clear that outside of 1AU, and especially for the Kuiper Belt this condition seems to be fulfilled.
	Exactly! That's the point we're getting at here. (TODO: do I need to clarify something in the paper here?)
263: Again, I think the concept of Hill radius and tidal stability needs to be formally introduced earlier.
	TODO: Introduce this formally earlier
273: Wouldn't these bodies then be tidally disrupted, and therefore not exist? Alpha > 1 is certainly not true for any solar system bodies that I am aware of. I understand that the scaling of alpha predicts alpha >1 at some radius (which I would explicitly mention), but this just implies that inside this radius, bodies must have greater densities to not be disrupted.
	Alpha = 1 doesn't correspond  to the roche limit (although they are close)! TODO: Maybe add a calculation here explaining this.
289: gravitational radius: is this physically motivated? Maybe reference
	There is a physical explanation of r_g immediately below equation 5 in the text. Also added a reference to Ida 1990.
296: beta > 1 means collision cross section > geometric and small gravitational focusing. I recommend explicitly stating what the physical intuition for beta < 1 is.
	Done.

@@

3. Numerical Methods
335: children: Would those not be the parent particles?

4. Narrow Annulus Simulations
352: I think it's somewhat ambitious calling four simulations a parameter study.
362: Regarding the initial condition, I recommend adding some more intuitive mass units, i.e. approx 0.5 Ceres masses, as well as the corresponding radius 341 km. Also, to what is the bulk density reduced to? Is this still physical? What is the corresponding radius? Maybe motivate the initial planetesimal masses / radii (e.g., Morbidelli 2009, Hartlep & Cuzzi 2020, Gerbig & Li, 2023). Also, maybe put the total mass of the planetesimals into context. I think this is 0.33 Earth masses, within 0.1 AU of Earth, the feeding zone of Earth.
367: I'm assuming the semi-major axis is also drawn from a uniform distribution?
377: is alpha=1 chosen because it is the highest stable alpha? If so I would write that. Why is beta = 15000 chosen?

Figure.2: Some suggestions to improve this figure are:
A reference planetesimal size that shows what mass a circle has
Horizontal lines in each panel showing where e_h is
Maybe consider having beta value as a column label and alpha value as a row label. I would also explicitly state the values for alpha.
I would also probably label the four panels / cases a,b,c,d and then refer as such in the text body

391: Why would this because there are no high eccentricity planetesimals. Isn't the fact that there is no residual population not already evidence that encounters are accreting rather than scattering? Also, isn't this also not just very expected since alpha exactly quantifies the relative importance of scattering vs accretion?
	There are two possible outcomes from an encounter:
		Scattering, eccentricity increases, subsequent encounters more unlikely
		Collision -> merger -> eccentricity decreases due to collisional damping
	No high eccentricity particles suggests there are very few scattering events
394: small alpha, large beta doesn't look dynamically cool to me.
	TODO: Change the text to only reference the dynamical state of the protoplanets
400: Small alpha large beta does not look dynamically cooler than large alpha large beta to me. In fact, the opposite.
	You are correct! This is a mistake in the text. TODO: Fix
402: I don't really understand why small alpha small beta leads to so many dynamically hot bodies, both planetesimals and protoplanets. Maybe the authors can elaborate.

Figure 3:
Again, I really think labeling the cases would be helpful.
I would express the mass in units of e.g. Ceres
406: I'm not sure about "significantly" larger. The biggest body in large alpha, small beta is about a factor 3 more massive than in small alpha, small beta. This is only a factor of 1.4 in size.
415: the y-label in Fig.6 does not match the description in the text
Figure 4:
Why don't the simulations start all with m_max/m = 1?
I'd be interested in also having another panel that shows the evolution of just M_max and <m> (maybe in solid and dashed lines) for the four simulations. That would also underscore that the turnover is driven by an increase in <m> rather than m_max.

418: I would specifically state that for the large alpha cases runaway growth only occurs at the start, to avoid initial confusion that the slope turns around when the planetesimal supply is depleted.
435: Is this true? I was under the impression that the residual planetesimal population remains? I think this does not seem consistent with the rest of the work...?
	TODO: Change the wording here. Maybe this isnt true. Would the growth mode simply 'flatten' out here?
446: What exactly does alpha being non-negligible mean? Given that alpha is always less than unity, what alpha value marks the transition of the oligarchic growth regime vs, consumption regime? See major comment.

### 2023-03-10 main.tex
Add references requested by referee

Regardless of how the mm- to km-sized growth barriers are surmounted, gravity begins to dominate and aerodynamic gas drag 
plays an increasingly unimportant role beyond this size. During this phase, collision cross sections are enhanced as gravitational 
focusing \citep{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Because gravitational focusing 
becomes more effective as bodies grow larger, a period of runaway growth occurs \citep{wetherill89, kokubo96, barnes09} and a 
power law spectrum of masses develop. Eventually, the largest bodies (known as oligarchs) dynamically heat the remaining 
planetesimals, severely limiting further growth \citep{kokubo98}. The final outcome of this phase is a bimodal population of 
dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest 
that the asteroid belt, Kuiper belt and the Oort cloud are largely composed of the leftovers of this stage of planet formation (add reference).

@@

Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received most of the attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, is more difficult and expensive to model because there are too
many particles to directly track with traditional N-body codes, while
the gravitational influence of the few massive oligarchs that
form make statistical methods inappropriate because the highest mass bins contain only a few bodies. Due to computational expense, N-body simulations of
planetesimal accretion are usually modeled in a narrow
ring \citep{kokubo96, kokubo98}, and the results and timescales are then scaled to suit whatever relevant orbital period is being 
studied. N-body simulations of terrestrial planet assembly typically begin with a series of neatly spaced oligarchs, whose mass 
varies smoothly with orbital period. As we will show in this paper, oligarchic growth does not scale to arbitrarily short
orbital periods.

Given that Systems of Tightly-packed Inner Planets (STIPs) appear to
be a common outcome of planet formation \citep{lantham11, lissauer11, rowe14}, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures \citep{izidoro17, izidoro21}, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place, and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical initial conditions (fully formed, evenly spaced protoplanets, e.g. \citet{raymond06}) used in studies of terrestrial planet formation are actually appropriate for understanding STIPs.

@@

In the limit that $v_{esc} \gg v$, it can be shown that $dM/dt \propto
M^{4/3}$, which implies a runaway scenario where growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \citet{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

An important feature is missing from the model described above, which
limits its applicability at late times. Gravitational stirring, which converts
Keplerian shear into random motion, raises the typical encounter velocity
between planetesimals over time \citep{weidenschilling89, ida90} and diminishes the effectiveness of gravitational
focusing. As the mass spectrum of the system evolves away from uniformity, these velocity differences become even more
pronounced. As the system evolves, it tends toward a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random 
motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body 
relaxation time \citep{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be $\approx 10$ for a planetesimal disk. Despite
the fact that the behavior of gravitational stirring is well-described
by a two-body formula, \citep{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters, which 
requires a three-body formalism. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the 
Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the 
main pathway for energy exchange between planetesimals and growing protoplanets is unable to operate.

@@

We also introduce a second dimensionless quantity, which relates the physical size of the bodies to the velocity state of the system
\begin{equation}\label{eq:beta}
	\beta = \frac{r_{pl}}{r_{g}}.
\end{equation}
where $r_{g} = G m_{pl} / v^{2}$ is the gravitational radius of a body. Encounters between bodies inside of a distance of $r_{g}$ result in significant deflections of their trajectories. It should be noted that the gravitational focusing enhancement factor $v^{2}/v_{esc}^{2}$ is equal to 1 for $\beta = 1$. In the case where $r_{g}$ is smaller than the size of a planetesimal, the gravitational focusing enhancement factor will be between 0 and 1 and the collision cross section is mostly set by the geometric value. \textbf{For very large values of $r_{g}$ ($\beta \gg 1$), the effective collision cross section is almost entirely set by gravitational scattering.}

>>>>

Regardless of how the mm- to km-sized growth barriers are surmounted, gravity begins to dominate and aerodynamic gas drag 
plays an increasingly unimportant role beyond this size. During this phase, collision cross sections are enhanced as gravitational 
focusing \citep{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Because gravitational focusing 
becomes more effective as bodies grow larger, a period of runaway growth occurs \citep{wetherill89, kokubo96, barnes09} and a 
power law spectrum of masses develop. Eventually, the largest bodies (known as oligarchs) dynamically heat the remaining 
planetesimals, severely limiting further growth \citep{kokubo98}. The final outcome of this phase is a bimodal population of 
dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest 
that the asteroid belt \textbf{\citep{bottke05, morbidelli09}}, Kuiper belt \textbf{\citep{levison08, sheppard10}} and the Oort cloud \textbf{\citep{levison11}} are largely composed of the leftovers of this stage of planet formation.

@@

Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received most of the attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, is more difficult and expensive to model because there are too
many particles to directly track with traditional N-body codes, while
the gravitational interactions between the \textbf{few massive bodies produced by the runaway growth phase \citep{ida93, kokubo95, kokubo98} render statistical methods inappropriate}. Due to computational expense, N-body simulations of
planetesimal accretion are usually modeled in a narrow
ring \citep{kokubo96, kokubo98}, and the results and timescales are then scaled to suit whatever relevant orbital period is being 
studied. N-body simulations of terrestrial planet assembly typically begin with a series of neatly spaced oligarchs, whose mass 
varies smoothly with orbital period. As we will show in this paper, oligarchic growth does not scale to arbitrarily short
orbital periods.

Given that Systems of Tightly-packed Inner Planets (STIPs) appear to
be a common outcome of planet formation \citep{lantham11, lissauer11, rowe14}, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures \textbf{\citep{izidoro17, izidoro21}}, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place, and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical initial conditions (fully formed, evenly spaced protoplanets, e.g. \citet{raymond06}) used in studies of terrestrial planet formation are actually appropriate for understanding STIPs.

@@

In the limit that $v_{esc} \gg v$, it can be shown that $dM/dt \propto
M^{4/3}$, which implies a runaway scenario where growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \citet{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk \textbf{\citep{lissauer87}, although one should note that pebble accretion \citep{lambrechts12, lambrechts14, bitsch15} is a viable alternative scenario}. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

An important feature is missing from the model described above, which
limits its applicability at late times. Gravitational stirring, which converts
Keplerian shear into random motion, raises the typical encounter velocity
between planetesimals over time \citep{weidenschilling89, ida90} and diminishes the effectiveness of gravitational
focusing. As the mass spectrum of the system evolves away from uniformity, these velocity differences become even more
pronounced. As the system evolves, it tends toward a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random 
motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body 
relaxation time \citep{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be \textbf{$\approx 3$ for a planetesimal disk \citep{ida90, stewart00}}. Despite
the fact that the behavior of gravitational stirring is well-described
by a two-body formula, \citep{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters, which 
requires a three-body formalism. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the 
Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the 
main pathway for energy exchange between planetesimals and growing protoplanets is unable to operate.

@@

We also introduce a second dimensionless quantity, which relates the physical size of the bodies to the velocity state of the system
\begin{equation}\label{eq:beta}
	\beta = \frac{r_{pl}}{r_{g}}.
\end{equation}
where $r_{g} = G m_{pl} / v^{2}$ is the gravitational radius of a body (see eq 4.1 of \citet{ida90}). Encounters between bodies inside of a distance of $r_{g}$ result in significant deflections of their trajectories. It should be noted that the gravitational focusing enhancement factor $v^{2}/v_{esc}^{2}$ is equal to 1 for $\beta = 1$. In the case where $r_{g}$ is smaller than the size of a planetesimal, the gravitational focusing enhancement factor will be between 0 and 1 and the collision cross section is mostly set by the geometric value. \textbf{For very large values of $r_{g}$ ($\beta \gg 1$), the effective collision cross section is almost entirely set by gravitational scattering.}
